Mental simulation as Bayesian Quadrature

### 2013-05-29 man/nips-2013.tex
Create initial version from final project writeup
Add Tom's comments/changes

\title{Optimal strategies in mental rotation tasks}
\author{Jessica B.~Hamrick\\
  Department of Psychology\\
  University of California, Berkeley\\
  Berkeley, CA 94720\\
  \texttt{jhamrick@berkeley.edu}}

@@

@@

@@

There is a long and rich history of work on mental rotation beginning
with \cite{Shepard1971}, who presented participants with pairs of
images such as those shown in Figure \ref{fig:mental-rotation}. These
pairs three-dimensional objects could either be: the same object
rotated in plane (Figure \ref{fig:mental-rotation}A), the same object
rotated in depth (Figure \ref{fig:mental-rotation}B), or two different
objects (Figure \ref{fig:mental-rotation}C; in this case, the
``different'' objects had reflective symmetry, but no rotational
symmetry). Participants had to determine whether the two images were
of the same object or not, and \cite{Shepard1971} famously found that
response times for plane and depth rotations had a strong linear
relationship with the minimum angle of rotation between the two
objects. The conclusion was that participants were visually
``rotating'' the objects in their minds.

@@

We assume that mental imagery is indeed a visual process, and turn to
the question of how, and when, it is used. Previous work has examined
how people might use imagery \cite{Hegarty2004, Schwartz1999} to solve
reasoning problems, and when they might use imagery as opposed to a
symbolic rule \cite{Schwartz:1996uy}. People seem to use imagery
methodically, which begs the question: what is the method?

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/shepard-rotation.png}
  \caption{\textbf{Classic mental rotation task}. Participants in
    \cite{Shepard1971} saw stimuli such as these, and judged whether
    each pair of shapes was the same shape in two different
    orientations, or two different shapes. \textbf{A} shows a ``plane
    rotation'', \textbf{B} shows a ``depth rotation'', and \textbf{C}
    shows two distinct objects.}
  \label{fig:mental-rotation}
\end{figure}

@@

However, we cannot compute $p(X_b\vert X_a, R)$ directly. Instead, we
introduce a new variable $X_R$ denoting a mental image, which
approximates $RX_a$. The $X_R$ are generated sequentially by repeated
application of a function $\tau$:
\begin{align}
  X_R&=RX_a\nonumber \\
  &=\tau(X_{R-r}, r)\nonumber \\
  &=\tau(\tau(X_{R-2r}, r), r)\nonumber \\
  &\ldots{}\nonumber \\
  &=\tau^{(\frac{R}{r})}(X_a, r)
  \label{eq:tau}
\end{align} 
Where $r$ is a small angle, and $\tau^{(i)}$ indicates $i$ recursive
applications of $\tau$. Using this sequential function, we get:
\begin{align}
  p(X_a, X_b\ \vert \ h_1)&=\int_R \int_{X} p(X_b\vert X) p(X\vert X_a, R)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  x&= \int_R \int_X p(X_b\vert X)\delta(\tau^{(\frac{R}{r})}(X_a, r)-X)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R p(X_b\vert X_R)p(X_a)p(R)\ \mathrm{d}R
\end{align}

@@

This process of mental rotation (i.e., generating a single $X_R$ (as
in Equation \ref{eq:tau}) and then computing $S(X_b, X_R)$) is a
computationally demanding cognitive process. Thus, our goal is to
minimize the number of rotations while still obtaining an estimate of
$Z$ that is accurate enough to choose the correct hypothesis. With
this in mind, we now examine several approaches to estimating $Z$.

@@

>>>>

\title{Mental simulation as Bayesian Quadrature}
\author{Jessica B.~Hamrick and Thomas L.~Griffiths\\
  Department of Psychology\\
  University of California, Berkeley\\
  Berkeley, CA 94720\\
  \texttt{\{jhamrick,tom\_griffiths\}@berkeley.edu}}

@@

% I think the main thing would be doing something more sophisticated
% in the search, rather than using the gradient. A less sophisticated
% adaptive sampling method seems like it would be best, and using von
% Mises makes sense.

@@

% The introduction could be made more general -- start with the
% question of how a rational agent should use her computational
% resources, and then get onto mental simulation/rotation as a special
% case of that, rather than starting with the phenomena. I would make
% the big picture goals clear at the start, and then segue into the
% mystery of mental rotation -- how do we know which way to rotate?

@@

There is a long and rich history of work on mental rotation beginning
with \cite{Shepard1971}, who presented participants with pairs of
images such as those shown in Figure \ref{fig:mental-rotation}. These
pairs of three-dimensional objects could either be: the same object
rotated in plane (Figure \ref{fig:mental-rotation}A), the same object
rotated in depth (Figure \ref{fig:mental-rotation}B), or two different
objects (Figure \ref{fig:mental-rotation}C; in this case, the
``different'' objects had reflective symmetry, but no rotational
symmetry). Participants had to determine whether the two images were
of the same object or not, and \cite{Shepard1971} famously found that
response times for plane and depth rotations had a strong linear
relationship with the minimum angle of rotation between the two
objects. The conclusion was that participants were visually
``rotating'' the objects in their minds.

@@

We assume that mental imagery is indeed a visual process, and turn to
the question of how, and when, it is used. Previous work has examined
how people might use imagery \cite{Hegarty2004, Schwartz1999} to solve
reasoning problems, and when they might use imagery as opposed to a
symbolic rule \cite{Schwartz:1996uy}. People seem to use imagery
methodically, which raises the question: what is the method?

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/shepard-rotation.png}
  \caption{\textbf{Classic mental rotation task}. Participants in
    \cite{Shepard1971} saw stimuli such as these, and judged whether
    each pair of shapes was the same shape in two different
    orientations, or two different shapes. \textbf{A} shows a ``plane
    rotation'', \textbf{B} shows a ``depth rotation'', and \textbf{C}
    shows two distinct objects. \TODO{update this caption}}
  \label{fig:mental-rotation}
\end{figure}

@@

However, we cannot compute $p(X_b\vert X_a, R)$ directly. Instead, we
introduce a new variable $X_R$ denoting a mental image, which
approximates $RX_a$. The $X_R$ are generated sequentially by repeated
application of a function $\tau$:
\begin{align}
  X_R&=RX_a\nonumber \\
  &=\tau(X_{R-r}, r)\nonumber \\
  &=\tau(\tau(X_{R-2r}, r), r)\nonumber \\
  &\ldots{}\nonumber \\
  &=\tau^{(\frac{R}{r})}(X_a, r)
  \label{eq:tau}
\end{align} 
% I wasn't clear whether the step that happens in Equation 6 should be
% in the model, or in the algorithm. I would probably be inclined to
% define the model in full generality, and then bring up the
% increments as part of the algorithm.
Where $r$ is a small angle, and $\tau^{(i)}$ indicates $i$ recursive
applications of $\tau$. Using this sequential function, we get:
\begin{align}
  p(X_a, X_b\ \vert \ h_1)&=\int_R \int_{X} p(X_b\vert X) p(X\vert X_a, R)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  x&= \int_R \int_X p(X_b\vert X)\delta(\tau^{(\frac{R}{r})}(X_a, r)-X)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R p(X_b\vert X_R)p(X_a)p(R)\ \mathrm{d}R
\end{align}

@@

This process of mental rotation (i.e., generating a single $X_R$ (as
in Equation \ref{eq:tau}) and then computing $S(X_b, X_R)$) is
computationally demanding. Thus, our goal is to minimize the number of
rotations while still obtaining an estimate of $Z$ that is accurate
enough to choose the correct hypothesis. With this in mind, we now
examine several approaches to estimating $Z$.

@@

% I think it would be compelling to add a second row to Equation 5
% where you show the averages and s.e. at each rotation for many noisy
% runs of the algorithm (possibly on different stimuli). If you can
% show that the average is a linear function, in a way that looks like
% the classic behavioral data, that will be compelling.
